A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-01 tex/codatadp.tex
more

@@

@@

\newcommand\keyw[1]{\textbf{#1}}

For the purposes of this paper, a (co)datatype specification consists of
$k$~mutually recursive types which are either all datatypes or all codatatypes.
Each datatype $\delta$ is equipped by $m \ge 1$ constructors, and each
constructor for $\delta$ takes $n \ge 0$ arguments, each associated with a
selector, and returns a $\delta$ value. The argument types must be either
already existing types or be one of the newly introduced (co)datatypes. The
introduced names for the (co)datatypes, the constructors, and the selectors must
be distinct with already introduced ones and among themselves. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, C_{11}\;(s_{111}{:}\; \sigma_{111}) \,\ldots\, (s_{11n_{11}}{:}\; \sigma_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\vdots} \\ % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\mid}}\; C_{1m}\;(s_{1m1}{:}\; \sigma_{1m1}) \,\ldots\, (s_{1mn_{1m}}{:}\; \sigma_{1mn_{1m}}) \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-.5\jot]
  \keyw{and}\; \,\delta_k & =\, \ldots
\end{aligned}
\]

(datatypes or codatatypes -- no mixture)
    * list of n constructors (n >= 1)
    * each constructor has m arguments (m >= 0)
    * each argument has a sort, which may be

  * don't consider mixtures (co)datatypes (cf. Agda), nor polymorphic types, nor
    is it possible to do any nesting; e.g. this is outside our scope:

      datatype tree = Node Int (Array tree)

>>>>

  * consider a natural fragment---datatypes as supported in modern SMT solvers
    and the SMT-LIB 2 standard, and codatatypes as their duals
    * mutual recursion, but no polymorphism, nested recursion

@@

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall

\paragraph{Notations.}
Blah

@@

\newcommand\keyw[1]{\textbf{#1}}
\newcommand\const[1]{\textsf{#1}}
\newcommand\ty[1]{\mathit{#1}}

For the purposes of this paper, a (co)datatype specification consists of
$l$~mutually recursive types which are either all datatypes or all codatatypes.
Each datatype $\delta$ is equipped by $m \ge 1$ constructors, and each
constructor for $\delta$ takes $n \ge 0$ arguments, each associated with a
selector, and returns a $\delta$ value. The argument types must be either
already existing types or be one of the newly introduced (co)datatypes. The
introduced names for the (co)datatypes, the constructors, and the selectors must
be distinct with already introduced ones and among themselves. Schematically:
%
\[
\begin{aligned}[t]
(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, C_{11}\;(s_{111}{:}\; \sigma_{111}) \,\ldots\, (s_{11n_{11}}{:}\; \sigma_{11n_{11}}) & \\
           & {}\;\phantom{=\,}\;{} \smash{\,\,\vdots} \\ % {}\;\phantom{=}\llap{\ensuremath{\mid}}\; \cdots \\
           & {}\;\phantom{=\,}\llap{\ensuremath{\mid}}\; C_{1m}\;(s_{1m1}{:}\; \sigma_{1m1}) \,\ldots\, (s_{1mn_{1m}}{:}\; \sigma_{1mn_{1m}}) \\[-1\jot]
   \smash{\vdots\,\,\,} \\[-1\jot]
  \keyw{and}\; \,\delta_l & =\, \ldots
\end{aligned}
\]
%
with
$\const{C}_{ij} : \sigma_{ij1}\times\cdots\times\sigma_{ijk_{ij}} \to \delta_i$
and $\const{s}_{ijk} : \delta_i \to \sigma_{ijk}$.

Polymorphic types, nested (co)recursion, and datatype--codatatype mixtures fall
outside this fragment. For example, the specification
\[\begin{aligned}[t]
      \keyw{datatype}\;\, \ty{tree} \,=\, \const{Node}\;\ty{int}\;\ty{forest} \\[-\jot]
      \keyw{and}\;\, \ty{forest} \,=\, \const{FNil} \mid \const{FCons}\; \ty{tree}\; \ty{forest}
\end{aligned}
\]
is supported, whereas
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}(\alpha) & \,=\, \const{LNil} \mid \const{LCons}\; \alpha\; \ty{llist}(\alpha) \\
      \keyw{datatype}\;\, \ty{tree} & \,=\, \const{Node}\; \ty{int}\; \ty{llist}(\ty{tree})
\end{aligned}
\]
is beyond the fragment considered here.
% (although it makes sense and is supported by Isabelle \cite{xxx}).
